Assessing Power of the California Microhap Baseline for GSI (and calculating Fst)

### 2024-03-15 002-GSI_and_Fst.Rmd
Trying some GenePop Stuff, but bailing on it now

```{r input_bypass, include=NEG_USING_SMK, eval=NEG_USING_SMK}
# inputs:
input_list <- list(
  final_baseline = "data/cali-chinook-baseline.rds",
  pop_labels = "inputs/reference-collection-names.csv",
  map_notations = "inputs/map-notations.tsv"
)
# outputs:
output_list <- list(
  assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
  fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
  assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
  assig_table_lfar_no_miss = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar-and-everyone-has-the-lfar.pdf",
  tex_table = "tex/inputs/samples-table.tex",
  gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
  gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
  gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf"
)
# you can create the necessary output directories like this:
dump <- lapply(output_list, function(x)
  dir.create(dirname(x), recursive = TRUE, showWarnings = FALSE)
)
```

@@

# get the full baseline with the new markers, etc.  We will toss out some
# non-winter Sacto fish of unknown provenance
full_base <- read_rds(input_list$final_baseline) %>%
  filter(collection != "SacNonW")

@@

# rename those collections and repunits
full_base2 <- full_base %>%
  rename(old_name = collection) %>%
  select(-repunit) %>%
  left_join(pop_labels %>% select(-run_timing_group), by = join_by(old_name)) %>%
  select(-old_name) %>%
  select(indiv, repunit, collection, sample_type, everything())

@@

# then extract bits of information from the data and set sparkline paths
fbmet2 <- fbmet %>%
  mutate(
    N = map_int(.x = data, .f = function(x) nrow(x)),
    `Run Timing` = map_chr(.x = data, .f = function(x) x$run_timing[1]),
    min_year = map_int(.x = data, .f = function(x) base::min(x$year, na.rm = TRUE)),
    max_year = map_int(.x = data, .f = function(x) base::max(x$year, na.rm = TRUE)),
    sparkline_path = str_c("tex/images/months-", abbrv, "_", str_replace_all(repository_name, " ", "-"), ".pdf")
  )

@@

# and now do some formatting
fbmet5 <- fbmet4 %>%
  ungroup() %>%
  mutate(
    `Sampling Months` = sprintf("\\raisebox{-0.12 em}{\\includegraphics[height=1.02em]{../%s}}", sparkline_path),
    `Year Range` = sprintf("%d--%d", min_year, max_year)
  ) %>%
  select(`Mainstem River`, abbrv, map_text, run_timing_for_table, ESU, N, `Sampling Months`, `Year Range`) %>%
  rename(
    `Mainstem River$^1$` = `Mainstem River`,
    Code = abbrv,
    `Location Name$^2$` = map_text,
    Run = run_timing_for_table,
    `ESU$^3$` = ESU,
    `$N$` = N
  )

@@

Now that we have the `upper_tri_fst` values in the right format, we just need
to add the WGS values to the lower triangle part of the matrix.

# Adding the WGS Fst values and making a figure

```{r}
# for now I am just going to put the WGS Fst's into the lower diagonal, but I think it will be better, ultimately,
# to but the bootstrapped CI's in there.
wgs_mat <- read_rds("../anderson-et-al-chinook-ecotypic-genetic-architecture/outputs/001/full-wgs-fst-mat.rds")

@@

# and then save that
ggsave(gsi_fst_fig_for_paper, file = output_list$gsi_fst_fig, width = 7, height = 14)

>>>>

```{r input_bypass, include=NEG_USING_SMK, eval=NEG_USING_SMK}
# inputs:
input_list <- list(
  #final_baseline = "data/cali-chinook-baseline.rds",
  final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
  pop_labels = "inputs/reference-collection-names.csv",
  map_notations = "inputs/map-notations.tsv"
)
# outputs:
output_list <- list(
  assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
  fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
  assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
  assig_table_lfar_no_miss = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar-and-everyone-has-the-lfar.pdf",
  tex_table = "tex/inputs/samples-table.tex",
  gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
  gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
  gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf"
)
# you can create the necessary output directories like this:
dump <- lapply(output_list, function(x)
  dir.create(dirname(x), recursive = TRUE, showWarnings = FALSE)
)
```

@@

# get the full baseline with the new markers, etc.  We will toss out some
# non-winter Sacto fish of unknown provenance
full_base <- read_csv(input_list$final_baseline) %>%
  filter(collection != "SacNonW")

@@

# rename those collections and repunits
full_base2 <- full_base %>%
  rename(old_name = collection) %>%
  select(-repunit) %>%
  left_join(pop_labels %>% select(old_name, collection, repunit), by = join_by(old_name)) %>%
  select(-old_name) %>%
  select(indiv, repunit, collection, sample_type, everything()) %>%
  filter(!duplicated(indiv))  # necessary cuz clemento stuck mill and deer together into a single code

@@

# Remove this once I get the full meta data
fbmet <- fbmet %>% filter(!is.na(WATERSHED))

# then extract bits of information from the data and set sparkline paths
fbmet2 <- fbmet %>%
  mutate(
    N = map_int(.x = data, .f = function(x) nrow(x)),
    N_date = map_int(.x = data, .f = function(x) sum(!(is.na(x$collection_date)) & yday(x$collection_date) != 1) ),  # we use YYYY-01-01 to denote year is known but day is missing in the data repository, so here we filter those out.
    `Run Timing` = map_chr(.x = data, .f = function(x) x$run_timing[1]),
    min_year = map_int(.x = data, .f = function(x) base::min(x$year, na.rm = TRUE)),
    max_year = map_int(.x = data, .f = function(x) base::max(x$year, na.rm = TRUE)),
    sparkline_path = str_c("tex/images/months-", abbrv, "_", str_replace_all(repository_name, " ", "-"), ".pdf")
  )

@@

# and now do some formatting
fbmet5 <- fbmet4 %>%
  ungroup() %>%
  mutate(
    `Sampling Months` = sprintf("\\raisebox{-0.12 em}{\\includegraphics[height=1.02em]{../%s}}", sparkline_path),
    `Year Range` = sprintf("%d--%d", min_year, max_year)
  ) %>%
  select(`Mainstem River`, abbrv, map_text, run_timing_for_table, ESU, N, N_date, `Sampling Months`, `Year Range`) %>%
  mutate(map_text = str_replace(map_text, ",.*$", "")) %>% # this removes the abbreviations after the commas on the map names
  mutate(N_date = replace_na(N_date, 0L)) %>%
  rename(
    `Mainstem River$^1$` = `Mainstem River`,
    Code = abbrv,
    `Location Name$^2$` = map_text,
    Run = run_timing_for_table,
    `ESU$^3$` = ESU,
    `$N$` = N,
    `$N_\\mathrm{day}$` = N_date
  ) %>%
  arrange(factor(Code, levels = collection_order))

@@

Now that we have the `upper_tri_fst` values in the right format, we just need
to add values to the lower triangle part of the matrix.

# Adding genic differentiation values to lower triangle and making a figure

We are going to use genepop to compute p-values for differentiation.  To do
that we need to make a genepop file.
```{r}
gpt <- full_base2 %>%
  select(-repunit, -sample_type) %>%
  pivot_longer(
    cols = -c(indiv, collection),
    names_to = c("locus", "gene_copy"),
    values_to = "allele",
    names_pattern = "(.*)[_.]([12])$"
  ) %>%
  group_by(locus) %>%
  mutate(alle_int = replace_na(as.integer(factor(allele)), 0L)) %>%
  ungroup() %>%
  mutate(alle_str = sprintf("%02d", alle_int)) %>%
  select(-allele, -alle_int) %>%
  pivot_wider(
    names_from = gene_copy,
    values_from = alle_str
  ) %>%
  mutate(geno = str_c(`1`, `2`)) %>%
  select(-`1`, -`2`) %>%
  pivot_wider(
    names_from = locus,
    values_from = geno
  ) %>%
  select(-indiv)

gpt_nest = gpt %>%
  mutate(collection2 = str_c(collection, ","), .after = collection) %>%
  group_by(collection) %>%
  nest()

# now we write that out to a file
outf <- "results/GSI_and_Fst/genepop.txt"
cat("Baseline data set in genepop format\n    ", file = outf)
cat(names(gpt)[-1], sep = ",", file = outf, append = TRUE)
cat("\n", file = outf, append = TRUE)
for(x in gpt_nest$data) {
    cat("POP\n", file = outf, append = TRUE); 
    write_tsv(x, col_names = FALSE, file = outf, append = TRUE)
  }
```

Now, we could use the genepop package to get results from that file, but it
is crazy slow, so we do not do this.
```r
td_file <- "results/GSI_and_Fst/genic_tests.txt"
genepop::test_diff(inputFile = outf, pairs = TRUE, outputFile = td_file)
```

OK! HERE IS WHAT WE WILL DO: MAKE THIS PARALLELIZABLE.  WE WILL MODIFY THE ABOVE
TO MAKE A GENEPOP FILE FOR EVERY LOCUS, AND THEN WE WILL JUST RUN OVER ALL THOSE
LOCI USING mclappy ON 8 CORES ON MY LAPTOP, TO MAKE IT FASTER.  THEN, WE WILL PRESENT
THE RESULTS AS A TINY-MULTIPLE HISTOGRAM OF THE P-VALUES IN EACH CELL.  KINDA COOL!

Making a figure.
```{r}

@@

# and then save that
ggsave(gsi_fst_fig_for_paper, file = output_list$gsi_fst_fig, width = 10, height = 19)
